---
title: "ClimateTestLab"
output: html_document
date: "2025-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Last week we learned about how we avoided disaster, even as the world
added 4 billion people in 50 years. The immediate danger of running
out of resources and food was mostly averted, due in large part to the
innovations of those additional people. However, a by-product of 
sustained economic and technological progress has been an
ever-increasing amount of CO2 being released into the atmosphere. In
the article we read this week, David Lam calculates what the future of
these emissions looks like given a set of assumptions for economic
growth, population growth, and the carbon intensity of the economy.
In the first part of the lab, we'll reproduce Lam's result. Then,
you'll have a chance to modify the assumptions and explore
alternatives for how the world may look in 60 years.


Some goals for this lab:

1. Understand how emissions could decrease as the world population continues to grow.

2. Learn how to model the effect of a variable by holding it constant in a simulation. 

3. Understand the assumptions underlying Lam's model. 

and seperately, 

4. Use the Malthusian harvest model to look at the optimum harvest rate.

## Lam's Approach

At the heart of Lam's approach is a decomposition of global carbon
emissions. 

He divides the world into sets of countries by income level. Thus his
additive decomposition is:
$$
\mathbf{C}(t) = \sum_i C_i(t).
$$
where $\mathbf{C}$ is global emissions (in tons) and $i$ indexes the
groups of countries:  low, low middle, upper middle,
and high income (LICs, LMICs, UMICs and HICs respectively),

For each group of countries, Lam separates
the effects of population, economy, and technology using the identity:
$$
\mbox{Tons of carbon emissions} = \mbox{# of people} \times {\mbox{$ of output}
\over \mbox{person}} \times {\mbox{Tons of Carbon emissions} \over
\mbox{$ of output}}
$$

In symbols, this is 

$$ C_i(t) = N_i(t) \times y_i(t) \times c_i(t) $$

where $C_i(t) is the tons of carbon emitted per year by countries in
group $i$ in year $t$, $N_i(t)$ is the number of people in countries
each group at time $t$, $y_i(t)$ is the per capita output in each
group at time $t$, and $c_i(t)$ is emissions per unit of output in
each group at time $t$.

To forecast emissions, you thus need to know future values of the
population, output per person, and carbon intensity.

For population, Lam uses the UN projections. Population is the most
certain of these forecasts since some of the people who will be alive
in 60 years have already been born. Rates of growth are highest for
the poorest countries -- and interestingly are actually lowest for the
next-to-highest income groups.

Here are the populations and population growth rate Lam uses:
```{r}
million = 10^6
N_2024.vec = c("lic" = 649, "lmic" = 3288, "umic" = 2792, "hic" = 1233)*million
R.vec = c("lic" = 0.0170, "lmic" = 0.0061, "umic" =  -0.0035, "hic" = 0.0000)

pop_table = rbind(N_2024.vec, R.vec)
## format first row with commas
pop_table[1,] <- formatC(pop_table[1,], format = "f", big.mark = ",", digits = 0)

print(pop_table)

```


For economic growth, he predicts higher yearly growth for lower income
countries than higher income ones. This is similar to the idea of
convergence in economics, that sees countries with less capital per
worker grow quicker because of their increased marginal product of
capital, with capital invested there seeing higher returns. While this
concept is debated, Lam uses the World Bank estimates, consistent
with the convergence prediction.

Here are per capita income and growth rates in per capita income rates
that Lam uses:

```{r}
y_2024.vec = c("lic" = 726, "lmic" = 2521, "umic" = 10528, "hic" = 52178)
g.vec = c("lic" = 0.030, "lmic" = 0.025, "umic" =  0.020, "hic" = 0.005)
econ_table = rbind(y_2024.vec, g.vec)
print(econ_table)
```

Forecasts of carbon intensity is quite uncertain. Lam relies on
estimates from the IPCC. He has the biggest reductions per-capita occurring in
the highest and lowest income countries. For the lowest income countries, this
is expected from technology "leap-frogging", or the LICs being able to move directly to greener technologies instead of following the same large increases seen in countries like the US or China.

Here is the carbon intensity and exponential rates of change that Lam
uses:

```{r}
#CO2/dollar 2024
c_2024.vec = c("lic" = 2/10000, "lmic" = 5/10000, "umic" = 5/7500, "hic" = 2/10000)
per_capita_reductions_2084.vec = c("lic" = 0.49, "lmic" = 0.76, "umic" = 0.6, "hic" = 0.27)

t=60


#CO2/dollar 2084
N_2084.vec = N_2024.vec * exp(R.vec * t)
y_2084.vec = y_2024.vec * exp(g.vec * t)

c_2084.vec = per_capita_reductions_2084.vec * c_2024.vec * N_2024.vec * y_2024.vec / (N_2024.vec * y_2084.vec)

#Carbon Intensity Reduction Rates
sp.vec = log(c_2084.vec/c_2024.vec)/60



#c_2084.vec = c_2024.vec * exp(s.vec * t)

#c_2084.vec = 
#s.vec = c("lic" = -.042, "lmic" = -.03, "umic" = -.029, "hic" = -.027)
#s.vec = 
emissions_table = rbind(c_2024.vec, sp.vec)
print(emissions_table)
#exp(60*s.vec)
```

You can see that carbon intensity (per $) is higher in poorer than in
richer countries, but that the rate of de-intensification is expected
to be faster in poorer countries than in richer ones.


## Lam's Result


Let's plug the quanties we have into the identity that Lam is using
and see if we can replicate the result he shows in figure 10 for CO2.


First, let's compute global carbon emissions in 2024

```{r}

C_2024.vec = N_2024.vec * y_2024.vec * c_2024.vec
print(C_2024.vec)

C_2024_world = sum(C_2024.vec)

## we should get 30-something billion tons per year
billion = 10^9
C_2024_world/billion
```


Now let's compute the same for 2084 (60 years into the future)

```{r}
t = 60
N_2084.vec = N_2024.vec * exp(R.vec * t)
y_2084.vec = y_2024.vec * exp(g.vec * t)
c_2084.vec = c_2024.vec * exp(sp.vec * t)


C_2084.vec = N_2084.vec * y_2084.vec * c_2084.vec
C_2084_world = sum(C_2084.vec)
carbon_world_ratio_2084_to_2024 = C_2084_world/C_2024_world

```
Bingo, we see that carbon emissions are about 52 percent less.

Now let's look at the counter-factual of what we think would have happened without population growth

```{r}
counterfactual_R.vec = c("lic" = 0,"lmic" = 0,"umic" = 0,"hic" = 0) ## assume no growth
counterfactual_N_2084.vec = N_2024.vec * exp(t * counterfactual_R.vec) ## we could have just assigned 2024 values but we illustrate using changing growth rates so you have a code template to explore other scenarios.

counterfactual_C_2084.vec = counterfactual_N_2084.vec * y_2084.vec * c_2084.vec
counterfactual_C_2084_world = sum(counterfactual_C_2084.vec)

counterfactual_carbon_world_ratio_2084_to_2024 = counterfactual_C_2084_world/C_2024_world
print(counterfactual_carbon_world_ratio_2084_to_2024)
```

```{r}
factual_N_2084.vec = N_2024.vec * exp(t * R.vec) ## Using real pop growth rates

factual_C_2084.vec = factual_N_2084.vec * y_2084.vec * c_2084.vec
factual_C_2084_world = sum(factual_C_2084.vec)

factual_carbon_world_ratio_2084_to_2024 = factual_C_2084_world/C_2024_world
print(factual_carbon_world_ratio_2084_to_2024)
```

Let's graph figure 3. (like Lam, with his colors)

```{r}
barplot(c(0, 100*sum(factual_N_2084.vec - counterfactual_N_2084.vec)/sum(counterfactual_N_2084.vec),100*(counterfactual_carbon_world_ratio_2084_to_2024-1), 100*(factual_carbon_world_ratio_2084_to_2024-1)), col=c("red","blue"), xlab="Population                                                      CO2", ylim=c(-60,40), ylab="Percent Change")
text(c(3.1,4.23),c(-55,-55), 100*round(c(counterfactual_carbon_world_ratio_2084_to_2024-1, factual_carbon_world_ratio_2084_to_2024-1), 3))
title("Projected Change in Population and CO2 2024-2084")
grid(nx = NA,
     ny = NULL)
legend(0.1,-30, legend=c("No Population Change", "Population Change"),fill=c("red", "blue"))
```



It's almost exactly the same!!

How can this be? Look by income group

```{r}

C_table = rbind("2024" = C_2024.vec,
      "no pop growth" = counterfactual_C_2084.vec,
      "w/ pop growth" = factual_C_2084.vec)
## hard to read so let's divide by billion
round(C_table / billion, 2)

```


So what appears to be the case is that pop growth is increasing
emissions in the lowest income countries by about 0.5 billion tons and
in the lower middle income countries by about 1.3 billion tons. But
because population is expected to decline in the upper middle income
countires, allowing for population change actually decreases emissions
by about 1.6 billion tons in this group. So the offsetting effects of
population in the poorer and upper middle income cancel out.  The
other thing we see in this table is that the big declines take place
in the two richer groups of countries whether one allows ppulation
change or not, for the simple reason that there's no population
_increase_ expected in these richer countries.

One way to see all of this is to note that the exponential rates all
multiply 

So we could rewrite the identity as

$$ C_i(t) = N_i(t) \times y_i(t) \times c_i(t) $$
$$C_i(t) =N_i(0) e^{r*t} \times y_i(0)e^{g*t}\times c_i(0)e^{s*t}$$
$$
C_i(t) = C_i(0) \exp{ t (r + g + s)}
$$

The sum of the growth rates tells us about growth rate in emissions
for each income-group.

```{r}

C_growth_rate_table = cbind(R.vec, g.vec, sp.vec, "sum of all (forecast)" = round(R.vec + g.vec + sp.vec, 3),
                          "sum w/o pop (counterfactual)" = g.vec + sp.vec)
print(C_growth_rate_table)
```


You can see in the last column that the growth rate of carbon emissions would have been
negative everywhere without population change. With population change,
carbon emissions will increase in the poorer countries, but from a
very low base. They will decrease as fast or faster in the rich countries.

# Graded questions

1. What if there were no "leap-frogging" in technology, so that all
   countries had the slowest improvement in carbon intensity -- that
   of the high income countries? (based on the values in sp.vec) Would
   global carbon emissions still decline by 2084? By how much? Comment 
   on your answer by indicating whether you were surprised or not and 
   offering an explanation in just a sentence or two.
   
2. What if rates of carbon intensity reduction were much slower than
   Lam assumes -- say 1/2 of his values? Would global carbon emissions
   still decline by 2084?
   
3. Nordhaus argues that there will be a Tragedy of the Commons if each
   country is allowed to make their own independent judgement about how
   much carbon to emit. One country, or group of countries could
   decide not to cut emissions in order to maintain a higher
   productivity. Lets say a tragedy of the commons does take place,
   and one income group decides not to cut emissions at all. In which
   group would this have the largest increase on future C02?
   
4. Multiple Choice

a.Which country group sees the largest percent increase in population?
    a. HIC
    b. UMIC
    c. LMIC
    d. LIC
    
b.Which country group sees the largest percent increase in output?
    a. HIC
    b. UMIC
    c. LMIC
    d. LIC
    
c.Which country group sees the largest reduction in total CO2 emitted?
    a. HIC
    b. UMIC
    c. LMIC
    d. LIC
    


Optimal Harvest Questions

5. For the following questions, use the "app" available
   http://shiny.demog.berkeley.edu/josh/harvest/ to gather data on the
   relationship between harvest rate "h" and the sustainable
   (steady-state) yield. You can do this by moving the slider from 0
   upwards by increments of 0.02.

a. There is a slight bug in the app, so that it reports sustainable
   yields even when harvest rate is unsustainable. How can you inspect
   the graphs to check on sustainability?

b. When h = 0.01, the sustainable yield is 0.0678, what is the
   sustainable yield when h = 0.02 and h = 0.04?

c. What harvest rate maximizes the sustainable yield? [Hint: try all
   of the slider values]

d. What harvest rate maximizes the yield during the first year of
   harvesting (this is about year 10 on the slider, or year 0 on the
   time axes of the graph)? (You can either report the value you find
   with the "app" or your theoretical answer)

e. Would harvest yield be higher in year "1" if we harvested at the
sustainable rate found in part 4.c than  or if we harvested .02 more
than the sustainable rate?


Congratulations! You've finished Lab 6. Whew!