---
title: "Econ/Demog C175 Lab 4: Population Aging -- Are we having too
few children?"
output: html_notebook
---

# Overview 

We saw earlier in our analysis of the Solow model
that slower popualtion growth was good because it allowed more capital
per worker, without having to increase savings rates. With the Solow
model, the lower the population growth rate the better. 

On the other hand, we know a slower growing (or shrinking) population
will also have an older age-structure, increasing the need to support
the elderly, but reducing the share of resources that needed to
support children. 

Lee et al. in their Science paper show that these forces counteract
each other. A conclusion of their modeling is that there is an optimum
population growth rate, which has produces just the right age
structure and capital-work ratio.

In this lab, we reproduce their findings for Japan. For the graded
portion, you will show what happens if we have a different survival
schedule, with more people living to older ages.



The lab is organized as follows:

1. Review of how birth rates translate to growth rates and population
   age structure
   
2. Looking at how age-structures influence aggregate production and
   consumption and which fertility rate allows for the highest
   per-capita consumption (without considering capital dilution or
   deepening). 
   
3. Including capital using the Solow model. 

4. The graded portion, in which you redo the calculations with higher
   survival probabilies and discuss your results.
   
Note: Answers and explanations to non-graded questions are in our bCourses site
under folder name "Lab."

```{r}
# Load in packages
# install.packages("data.table")
library(data.table)
```

# 1. Fertility and age structure


## A. From Kids to Population Growth Rates

We begin by reviewing the relationship between the number of children
that people have and the population growth rate. Defining the Total
Fertility Rate (TFR) as the number of children born on average to each
woman, then the ratio of one generation to the next known at the "Net
Reproductionn Rate" or "NRR") is about
$$
NRR \approx TFR * \ell(\mu) * p_f
$$ 
where $\ell(\mu)$ is the fraction of children who survive to the mean
age of childbearing -- usually about 30 -- and $p_f$ is the proportion of births that are
female.

For example, if the TFR is 3.0 children per woman and 98 percent of
children survive to age 30, then

```{r}
TFR = 3
l_mu = 0.98
frac_female = 0.4886
NRR = TFR * l_mu * frac_female
print(NRR)
```

You should get a number like 1.43.

To convert this to an annual growth rate, we use the same approach as
we did for calculation exponential growth rates from populations at
two different points in time. If we started with 100 people in year "0",
then with this generational growth rate we would have 143 people in
year "30". So,
$$
r \approx [\log(NRR) - \log(1)] /[\mu - 0] = \log(NRR)/\mu.
$$

Continuing with our example,
```{r}
mu = 30
r = log(NRR)/mu
print(r)
```
gives us a population growth rate of about 1.2 percent.


What is the population growth rate implied by a long-term growth Total
Fertility Rate of 1.75 children per woman, assuming the same survival
and proportion born female as we did above?

A. About half-a-percent
B. About zero
C. About negative half-a-percent
D. About negative one-percent

Hint: change the TFR value in the first block of code

## B. From Population Growth Rates to Stable Age Structures


If demographic rates are constant for a long time, then the
age-structure will settle down to a steady state, with constant
proportions of people at each age. That is, the popualtion as a whole
may be growing or shrinking but the age-distribution will be constant
in terms of proportions at each age.

The formula for the stable age-structure is 
$$
n(x) = {e^{-rx} \ell(x) \over
\sum e^{-rx} \ell(x) },
$$
where $r$ is the population growth rate, $\ell(x)$ is the fraction
that survive to age $x$, and $n(x)$ is the proportion of the
population aged $x$.

We can look at the stable age-structures of a population that has
the survival schedules of contemporary Japan with positive, zero, and
negative growth rates.

```{r}
## use Japanese both_sex survivor rates
url <- "https://raw.githubusercontent.com/josh-goldstein-git/econ_demog_c175/main/data/JPN.bltper_1x1.txt"
dt <- read.table(url, header = TRUE, skip = 2)
#lx <- dt[dt$Year == 2021, "lx"] / 10^5
lx <- dt[dt$Year == 2009, "lx"] / 10^5
```


```{r}
lx <- lx
r.grow = +1/100
r.stationary = 0
r.shrink = -1/100
r.vec <- c(r.shrink, r.stationary, r.grow)
x <- 0:110
par(mfrow = c(1,3))
for (i in 1:3)
{
    r <- r.vec[i]
    nx <- exp(-r * x) * lx  / sum(exp(-r * x) * lx)
    plot(nx, x, type = 'n')
    segments(x0 = 0, x1 = nx, y0 = x)## barplot(nx, at = x, horiz = T)
    title(paste0("r = ", r))
}
```

Even with the realistic age schedule of survival, we see that the
stable age-structures look very much like the stylized pictures we saw
in class.

## 2. How age structures affects production and consumption

In Lee's set-up, each country has its own age-schedules of production
$y(x)$, which can be thought of as the dollar (or other currency)
value of what a person aged $x$ produces. Children produce nothing,
working-age adults produce more, and the elderly produce less. 

Each country also has its own age-schedule of consumption $c(x)$,
where consumption not only includes things like housing and food and
but also services like health care and education.

Let's look at the produciong and consumption schedules for Japan.

```{r}
## get the estimated age schedules of consumption and production
library(data.table)
dt <- read.table("https://raw.githubusercontent.com/josh-goldstein-git/econ_demog_c175_2023/main/data/japan_nta_schedules.txt")
head(dt)
#dt[, V1 := NULL]
cx <- dt$cx /150 ## converting yen to dollars
yx <- dt$yx /150
```

```{r}
x = 0:110
plot(x, yx, col = "blue", type = 'l', ylab = "Dollars per year",
     xlab = "age")
lines(x, cx, col = "red", type = 'l')
text(80, 20000, "Consumption c(x)", col = 'red')
text(80, 5000, "Production y(x)", col = 'blue')
title("Production and consumption schedules for Japan \n (in US dolllars")
```

Some Questions:

- Why is consumption higher for those aged 14-21 than for those aged
  21-30?
  
  A. Because education is counted as consumption.
  B. Because Nintendo game consumption is very expensive
  C. Because teenagers eat more.
  D. Because working people save for later.
  
  
- Why does consumption climb above age 60?

  A. Because the elderly are rich and spending their life-time savings
  on vacations and consumer goods.
  B. Because consumption includes healthcare, which rises very rapidly
  in cost as people get older. 
  C. Because the elderly are dying and the inheritances transferred to
  their children are counted as consumption.
  

- Does this graph mean that the elderly consume most of what society
  produces?
  
  A. Yes, because the curve is highest at these ages
  B. No, because this curve is per person alive at those ages, and
  there are fewer people alive at older ages.
  C. No, in order to find out the share of consumption at age $x$ you
  would want to weight the consumption schedule $c(x)$ by $n(x)$, the share of
  people aged $x$.
  D. Both B and C.
  

In order to measure the effect of age-structure on the aggregate
economy, we use Lee et al.'s measure of the "Support Ratio" (SR).  The
Support Ratio gives us an indicator the number of effective earners
per effective consumer in the population, based on age-specific
schedules of earnings $y(x)$ and consumption $c(x)$. 

With $N(x)$ people aged $x$, total production will be
$$
{\sum_x N(x) y(x)}
$$
and total consumption (for the baseline schedule) will be
$$
{\sum_x N(x) c(x)}
$$.

The ratio of these two is Lee's definition of the Support Ratio:
$$
SR = {\sum_x n(x) y(x) \over \sum_x n(x) c(x)}.
$$

If the Support Ratio is 2, that means that the economy is producing
twice what the average person in the baseline schedule would
consume, allowing a lot of consumption. If the support ratio is 0.6,
then the economy is producing only 60% of what the baseline consumer
would consume, implying less consumption than baseline. 

We are now -- at last -- ready to see how fertility affects the
Support Ratio! And maybe we can find the fertility rate that maximizes
it.

Let's start with a vector of total fertility rates and convert these
to population growth rates

```{r}
tfr.vec <- seq(1, 4, .1)
l_mu = 0.99 ## assume very low mortality up to age 30 (mu)
mu = 30
r.vec <- log(tfr.vec * 0.4886 * l_mu)/ mu
plot(tfr.vec, r.vec)
grid()
```

We see that a TFR of about 2 implies a long term growth rate of
about 0, whereas a TFR of 1.5 implies a long-term growth rate of about
-1 percent, and a TFR of 2.8 implies a long-term growth rate of about
1 percent.

Let's see how the SR varies with the growth rate $r$:


```{r}

sr.vec <- NULL
for (i in 1:length(r.vec))
{
    r <- r.vec[i]
    nx <- exp(-r * x) * lx  / sum(exp(-r * x) * lx)
    sr <- sum(nx*yx)/sum(nx*cx)
    sr.vec[i] <- sr
}
## print(round(sr.vec, 2))
par(mfrow = c(1,1), mar = c(5, 4, 4+2, 2) +.1)
plot(r.vec, sr.vec,  ylim = c(0, 1),
     ylab = "Support Ratio",
     xlab = "Population growth rate 'r'")
mtext("Total Fertility Rate", side = 3, line = 2.5)
title("Support Ratios vary with growth rates", line = 5)
## add an axis for TFR
axis(3, at = r.vec, labels = tfr.vec)
grid()
opt.tfr.transfers <- tfr.vec[which.max(sr.vec)]
abline(v = r.vec[which.max(sr.vec)])
par(mar = c(5, 4, 4+2-2, 2) +.1)
```

What can we conclude?

A. Without accounting for capital, the age-schedules of production and
consumption in Japan suggest that the optimum fertility rate would be
about 2.4 children per woman.

B. Actually, within a broad range of fertility from (about 1.5 to
3.5), age-structure doesn't appear to matter much, because the Support
Ratio isn't that much lower, even when it's not at the optimum.

C. The reason that the age-schedule doesn't matter that much is
because higher growth rates mean more kids and fewer elderly, whereas
lower growth rates mean fewer kids and more elderly -- and the two
changes in the different ages of dependency tend to cancel each other
out.

D. All of the above!!




## Part 2: Another Capital Idea

Of course, changing age structures are not the only effect of
different rates of population growth. The other effect that Lee et all
look at is the effect of the growth rate itself
on capital deepening or dilution.

Recall in the Solow model that higher population growth rates require
higher savings rates in order to keep per capita production constant.

Including this in our model should tend to lower optimal fertility,
since it provides an additional advantage to lower growth rates.


There are several possible ways to include capital in our
models. (Those interested in the details can consult the
"supplementary materials" of the Lee et al paper in Science).  For the
lab, we will reproduce the "low cost capital approach," which assumes
that the capital-output ratio stays constant over time and that
savings adjusts to maintain capital per worker.

In the Solow model, we saw early in the semester that if one wants to
preserve a constant stock of capital per person, faster population
growth rates need to be matched with larger fractions of income saved.

In the steady state, we have savings per person equaling the new
capital needed to keep up with population growth $r$, depreciation
$d$, and technological advances $\lambda$,
$$
s y = (r + d + \lambda) k
$$

We can rearrange this steady-state to give us.

$$
s = (r + d + \lambda) (k / y)
$$

Lee et al.  assume standard values such that $k/y \approx 3.0$, 
$d= 0.05$, and $\lambda = .02$. This allows calculation of $s$ 
as a function only of $r$.

```{r}
d = .05
lambda = .02
ky.ratio = 3.0
s.vec = (r.vec + d + lambda) * ky.ratio
plot(r.vec, s.vec)
grid()
title("Savings needed for different population growth rates")
```
In the figure you can see that a population shrinking by 1 percent a
year requires about an 18 percent savings rate whereas the population
growing by 1 percen requires a 24 percent savings rate.


To take account of the effect of increased savings on the support
ratio, we reduce the Support Ratio by a factor of
$(1-s)$. This operationalizes the idea that the more savings is
required, the less production is available for consumption.

$$
SR_{capital} = (1- s_r) { \sum_x e^{-rx} \ell(x) y(x) \over \sum_x e^{-rx} \ell(x) c(x)},
$$
where we use the savings rate $s_r$ that corresponds to the
population growth rate $r$.

```{r}

## This is  the "low capital" case, where the capital-income ratio is
## fixed at 3.0. This is would happen if we had Cobb-Douglas.

## In this case, the a share "s" needs to be set aside from income as
## savings, and cannot be consumed. So we exclude this share from the
## numerator of the support ratio.

SR.low.capital.vec <- NULL
tfr.vec <- exp(r.vec*30)/0.4886
Lx <- lx

## let's get the savings rate for each growth rate
s.vec <- (r.vec + d + lambda) * ky.ratio
plot(tfr.vec, r.vec)
for (i in 1:length(r.vec))
{
    r <- r.vec[i]
    my.Nx <- Lx * exp(-r * x)
    my.s <- s.vec[i]
    SR.low.capital.vec[i] <-   (1-my.s) * sum( my.Nx * yx) / sum(my.Nx * cx)
}

par(mfrow = c(1,2))
## no capital case
plot(tfr.vec, sr.vec,  ylim = c(.5, .8),
     ylab = "Support Ratio",
     xlab = "TFR")
opt.tfr.no.capital <- tfr.vec[which.max(sr.vec)]
grid()
abline(v = opt.tfr.no.capital, lty = 2)
text(opt.tfr.no.capital, max(sr.vec), opt.tfr.no.capital, pos = 3)
title("Age-structure effects only")
## add an axis for TFR
plot(tfr.vec, SR.low.capital.vec, ylim = c(.5, .8) -.2,
     col = 'red',
     ylab = "Support Ratio",
     xlab = "TFR")
grid()
title("Age-structure and capital effects")
opt.tfr.low.capital <- tfr.vec[which.max(SR.low.capital.vec)]
print(opt.tfr.low.capital)
abline(v = opt.tfr.low.capital, col = "red", lty = 2)
text(opt.tfr.low.capital, max(SR.low.capital.vec), opt.tfr.low.capital, pos = 3,
     col = "red")
## plot(r.vec, SR.low.capital.vec, ylim = c(0, 1))
## grid()
## lines(r.vec, sr.vec)
```

## Graded Questions

>    1. Use the PAYGO identity to explain in a few sentences why
>       higher fertility makes it easier to finance social security.
> 	   
>    2. In the Science paper, Lee et al find that the optimal fertility
>       rates for richer countries are higher than for poorer
>       countries. Explain in a few sentences why this is.
> 	  
>    3. In the code for the lab, we used the same 2009 schedule of
>       mortality that Lee et al used in their calculations. Mortality
>       has declined in Japan quite a bit since then. What do you think
>       would happen to the level of optimal fertility if you used the
>       2021 mortatity rates, which would have more people surviving to
>       older ages? Explain your reasoning in a few sentences. (Note:
>       you don't have to do any calculations to answer this question --
>       we'll do that next :)
> 	  
>   4. Change the code above so that it uses the 2021 survival schedule
>      instead of 2009. Fill in the following cells of the table:
> 
>                    Table: Optimal fertility in Japan, by model and
>                    survival 

> 		                                 Survival schedules
>                   										 2009    2021
> 	 model:  age-schedules only          2.376     B.
> 	         age & capital               A.        C.
> 	  
> 	  (You should be able to get A from the lab without making any
>       changes. For B & C, you'll need to redo the calculations with
>       the new survival schedule).
>. 
>.   Describe in a sentence or two the effect of improving survival on optimal 
>    fertility based on the results you put in this table. 
>
>   5. Is the following True or False, and why? (explain in a sentence
>      or two)
>   
> 	  Increasing survival might change Lee's measure of optimal fertility from a TFR
>.    less than 2 to a TFR greater than 2.
> 	  
>   6. List at least one other consideration for optimal fertility that
>      is not accounted for in the analysis we did.
> 	 

Congratulations! You are finished with Lab 6.