---
title: 'Econ/Demog C175 Lab 5: Population and Climate Modeling'
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

This week's lab will use a simplified version of Nordhaus's climate
model. We will not get too much in to the details of the
model. Instead we will use it to focus on the potential 
effect of population growth on climate change.

Our questions will be:

1. How does the model specify the relationship between population and
climate?
   
2. What would happen if we slowed population growth? How big an effect
would it have on global temperature if we slowed growth by 50%? How
about if we just had zero population growth from now on?

3. Is part of the benefit of slower population growth offset by
   capital deepening?

4. How does the effect of slower population growth compare to an
   acceleration in decarbonization of the economy?


# The model

As discussed in the Nordhaus reading, Integrated Assessment Models are
used to try to understand relationships between economic activity and
climate change. The goal is not to get exact predictions but rather to
get a useful approximation of the magnitudes involved. The questions
can be what-if. For example, if we were able to reduce the emissions
by 10%, how would that change the climate, and how would that change
our standard of living? Or the questions can be more prescriptive. For
example, how high would carbon taxes need to be in order to maximize
the welfare of humanity over the next 2 centuries?

Today our goal is to focus mostly on population and climate, and so we
will leave many of the other interesting economic questions
aside. (These include: How does climate change feedback and hurt the
economy? How should one take into account the welfare of different
generations living at different points in time? How do we account for
risk-aversion, meaning the chance that things could get very, very
bad?).

We're going to use a simplified Integrated Assessment Model developed
by Panagiotis Tsigaris and Joel Wood, at Thomson Rivers University in
British Columbia. Special thanks to them for sharing their R code, a
very slightly adapted version of which we will use below.

i. A Solow-like production function, with time varying inputs.

$$
Y(t) = A(t) K(t)^a L(t)^{1-a}
$$


Population will be exogenous, but capital (per person) will depend on
the usual balance of savings, depreciation, and population
growth. $A(t)$ is a term representing productivity (or
"technology"). 

(There is also an additional term $D(t)$, which will appear in
iv. below, but which is left out for now to emphasize the similarity
to what we did with Solow.)


ii. Carbon emissions ($E(t)$) as a function of economic output

$$
E(t) = \sigma(t)  Y(t)
$$ 

Emissions are tons of carbon released. A crucial quantity here is
$\sigma(t)$, which is the carbon intensity of the economy, telling us
the average carbon emission per $ unit of output. Declining
$\sigma(t)$ means that the economy is "getting greener" in one
sense. But if the economy grows quickly, then the increase in output
$Y$ could outweigh the decrease in emissions intensity $\sigma$.

and

iii. Temperature change as a function of cumulative carbon emissions.

Cumulative emmisions are defined as

$$
C(t) = C_0 + \sum_0^t E(\tau).
$$

And temperature change is given by

$$
T(t) = \beta * C(t).
$$

The climate part of this model is very simple. It just says that
temperature deviation from normal rises linearly at rate $\beta$ with
emissions.

(Warning: I'm not a climate scientist and don't know how realistic
this is. But, as we will see, the projected temperature increase is
similar in this simplified model as it is in Nordhaus's more
sophisticated approach.)


iv. A "damage" function that makes the economy less productive as
temperature rises.

$$
Y(t | damage) = D(t)  Y(t)
$$ 

We're mostly going to ignore this for now, although it is in the code
and has some effect, especially over the very long term. Getting this
right is important for assessing the Social Cost of Carbon, and is a
very interesting scientific topic on its own in terms of studying the
impacts of warming.


# The R Code version of the model

Here is the full model, in R, as calibrated by Tsigaris and Wood to be
close to the estimates of more complicated models. You don't need to
understand the code in detail, just understand that it is a dynamic
model that begins with initial conditions and loops forward in
time. For our purposes, I've parameterized the function to allow
changes in population and in decarbonization. The default values are
given in the function definition, but can be changed as we do later in
the Lab.

(Note: the model begins in 2010, and so it gives us 13 years to
adapt that we no longer have. We could (but we have not) update
all of the initial conditions to be consistent with today.)

```{r}
library(data.table)

## if you're working on this using the RStudio desktop version, make
##   sure to install this when the app prompts you to - the last line
##    under the Jobs tab should be "✔ Package 'data.table' successfully
##   installed."
## you can also run install.packages("data.table")
   

solow_climate_fun <- function(gL0 = 0.022547, dL = -0.052,
                              gsigma0 = -.01, dsigma = -.0002)
{
    ## CONSTANTS ##
    ## capital's share of income
    a<-0.3
    ## savings rate
    s<-0.25
    ## Depreciation rate
    d<-0.1
    ## initial GDP, & Capital
    Y0<-63.69
    K0<-135
    ##Initial population, pop growth rate, and parameter reducing pop growth
    L0<-6.838 ## billions
    ## next lines commented out because they are arguments in fun
    ##  gL0<-0.022547 ## Initial growth rate, about R
    ##  dL<-0.052 ## exponential rate of decline in R, i.e,  R(t+1) = R(t) exp(-dL * 1)
    ## Initial Total Factor Productivity, TFP growth rate, and parameter affecting TFP growth
    A0<-3.955
    gA0<-0.015323
    dA<-0.011
    ## Initial emissions intensity, intensity growth, and parameter affecting intensity growth
    sigma0 <- 0.549  ## this is carbon output in some unit of tons per $ of output
    ## next lines commented out because they are arguments in fun
    ## gsigma0<- -0.01 ## initial rate of decline in sigma
    ## dsigma <- -0.0002 ## rate of change in rate of decline
    ## Damage Parameters ## we'll ignore these, but they do play as small role.
    theta1<-0.0023888849
    theta2<-2
    theta3<-0.00000507029
    theta4<-6.754
    ## Damage to TFP
    gamma<-0.001
    ## Damage through depreciation
    d1<-0.01 ## this is the same constant "d" we had in solow model
    ## CCR Cumulative Carbon
    beta<-1.8 ## multiplier to convert cumulative carbon to temperature increase
    C0<-530 ## total carbon output up to 2010

    ## ### BASE MODEL ###
    T<-191
    year<-c(2010:2200)

    ## intialization
    k<-rep(NaN,T) # rep means 'NaN' replication 'T' number of times
    A<-rep(NaN,T)
    gAt<-rep(NaN,T)
    sigma<-rep(NaN,T)
    gsigmat<-rep(NaN,T)
    y<-rep(NaN,T)
    D<-rep(NaN,T)
    L<-rep(NaN,T)
    gLt<-rep(NaN,T)
    E<-rep(NaN,T)
    Ccum<-rep(NaN,T)
    Temp<-rep(NaN,T)
    y.nd<-rep(NaN,T)
    k.nd<-rep(NaN,T)
    ## starting values (for notations see loop step, line 218 onwards)
    Ccum[1]<-C0
    Temp[1]<-beta*Ccum[1]/1000
    D[1]<-1/(1+theta1*(Temp[1]^theta2))
    A[1]<-A0
    gLt[1]<-gL0
    k[1]<-((s*A[1]*D[1])/(d+gL0))^(1/(1-a))
    gAt[1]<-gA0
    sigma[1]<-sigma0
    gsigmat[1]<-gsigma0
    y[1]<-D[1]*A[1]*k[1]^a
    L[1]<-L0
    gLt[1]<-gL0
    E[1]<-sigma[1]*y[1]*L[1]/3.67
    k.nd[1]<-((s*A[1])/(d+gL0))^(1/(1-a))
    y.nd[1]<-A[1]*k.nd[1]^a

    ## now we loop thru
    for (t in 2:T){
        ## with climate damages
        Ccum[t]<-Ccum[t-1]+E[t-1] ## cumulative emissions
        Temp[t]<-beta*Ccum[t]/1000 ## temperature increase in celsius
        D[t]<-1/(1+theta1*(Temp[t]^theta2)) ## "damages"; line 108
        gAt[t]<-gAt[t-1]/(1+dA) ## change in growth rate of TFP
        A[t]<-A[t-1]*(1+gAt[t]) ## TFP
        gLt[t]<-gLt[t-1]*(1+dL) ## change in growth rate of pop, dL < 0, gLt shrinks
        L[t]<-L[t-1]*(1+gLt[t]) ## Pop
        k[t]<-((s*A[t]*D[t])/(d+gLt[t]))^(1/(1-a)) ## capital per person in solow steady state
        gsigmat[t]<-gsigmat[t-1]/(1+dsigma) ## change in emissions intensity
        sigma[t]<-sigma[t-1]*(1+gsigmat[t]) ## emissions intensity
        y[t]<-D[t]*A[t]*k[t]^a ## income per capita
        E[t]<-sigma[t]*y[t]*L[t]/3.67 ## total emmissions (3.67 converts carbon to C02)

        ## without climate damages i.e., D=1
        k.nd[t]<-((s*A[t])/(d+gLt[t]))^(1/(1-a))
        y.nd[t]<-A[t]*k.nd[t]^a
    }
    out = data.table(year, Ccum, Temp, D,
                     A, L, k,
                     sigma, y, E, k.nd, y.nd)
    return(out)
}


```

# The baseline case, "Business As Usual"

```{r}
out = solow_climate_fun() ## without arguments, uses default
year = out$year
```

Let's walk through some of our exogenous inputs in the baseline scenario

* Population growth rates
```{r}
L = out$L
R = c(diff(log(L))/1, NA) ## calculate exponential growth rate as in Lab 1.
plot(year, R, ylab = "Crude Growth Rate",
     main = "Exponential Pop Growth Rate 'R'")
grid(lwd = 2) # lwd stands for line width
```
We see pop growth rate declines quickly to very low levels in 2050 and nearly zero by 2100.

* Productivity
```{r}
A = out$A
plot(year, A, ylab = "Multiplier",
     main = "Productivity factor")
grid(lwd = 2)
```
We see that productivity increases roughly linearly until about 2100,
after which it slows a bit.


* Emissions intensity
```{r}
sigma = out$sigma
plot(year, sigma, ylab = "Emissions per $",
     main = "Emissions intensity")
grid(lwd = 2)
```

Emissions decline fairly linearly, but rate of decline slows late in
projection.

Now we're ready to look at what will happen to the climate and the
economy. Before we begin, let's make some guesses:

Q1 Will per capita output $y$ in 2100 compared to 2010 output be:

A. A lot lower because of climate change damages.
B. The same because technology is assumed constant
C. Much higher (> 3x) because we assumed a lot of productivity
improvement, so much so that climate change damages won't make much of
a dent.
D. Impossible to say

Let's see:
```{r}
y = out$y
plot(year, y, ylab = "$ per person",
     main = "Output, per capita")
grid(lwd = 2)
```
It looks like the answer is "C"


Q2 How high will temperature rise?
A. Zero, because this is the so-called baseline case.
B. By 2 degrees Celsius because that is what is considered still
   not-catastrophic
C. By much more than 2 degrees, because baseline is business-as-usual,
and if we continue on the path we are on, there will be a lot of
warming.

Let's see:
```{r}
Temp = out$Temp
plot(year, Temp, ylab = "Degrees, Celsius",
     main = "Warming")
grid(lwd = 2)
```
It looks like the answer is, once again, "C".


Let's compare to the Nordhaus nobel prize
lecture.(https://www.nobelprize.org/uploads/2018/10/nordhaus-lecture.pdf)
In his Figure 5, on page 2003, you can see the "Temperature
trajectories for different objectives.The blue line shows "Base",
which has 2 degrees of warming by 2050 and 4 degrees by 2100. This is
very close to what the simple model we are using is predicting. (This
is probably not evidence that either model is right, but rather an
indication that the model defaults of our simple model were chosen so
that their output would be close to the standard model, Nordhaus's.)

# Slower population growth

As we saw above when we plotted population growth rates, the baseline
model assumes a leveling off of population growth, beginning at a bit
more than 2% a year and declining continuously to about zero
population growth by 2100.

Let's consider the case of speeding up the world demographic
transition so that the growth rate itself declines exponentially at
twice the rate. I call this the "slow pop" scenario to indicate that
population is growing more slowly.

```{r}
out.slow.pop <- solow_climate_fun(
	     gL0 = 0.022547, ## we leave starting pop growth as is
             dL =  -0.052*2, ## and double rate of decline in pop growth rate 
	     gsigma0 = -.01,
	     dsigma = -.0002)
```

Let's see how much difference this made in population.
```{r}
L.slow <- out.slow.pop$L 
plot(year, L, main = "Population")
lines(year, L.slow)
```
It looks like there would only about 8.3 billion people rather than
10.3 billion. This is a difference of about 20%.

Since the climate model depends on cumulative emissions and it doesn't
matter when the emissions occurred, it is useful to think about the
number of "person-years" in the two scenarios. To find the
person-years lived from 2010 to 2100, we add up the people alive in
each year and sum them up.

```{r}
pyl.base = sum(L[year %in% 2010:2100])
pyl.slow = sum(L.slow[year %in% 2010:2100])
print("base")
print(pyl.base)
print("slow")
print(pyl.slow)
print("pop years lived ratio")
print(pyl.slow / pyl.base)
```

It looks like person-years will decline by about 15%. How about
emissions?

```{r}
E = out$E
E.slow = out.slow.pop$E
sum.E.base = sum(E[year %in% 2010:2100])
sum.E.slow = sum(E.slow[year %in% 2010:2100])
print("E ratio")
print(sum.E.slow / sum.E.base)
```
Also, about 15%! (OK, 14% ...)

And now we can check on warming 

```{r} 

Temp.slow = out.slow.pop$Temp
plot(year, Temp) 
lines(year, Temp.slow) 
grid(lwd = 2) 

``` 
It looks like we have about +4 degrees in 2100 in the baseline, and about +3.6 degrees in the slower population growth scenario. This seems at first to be a decline of temperature increase by about 10%, since 3.6/4.0 = 0.9. But we started in 2010 already with +1 degree of warming. So the ratio we want is (3.6 -1)/(4 - 1) = 0.87 -- a 13% decline, very close to the 14% above.

Note that by 2200 the smaller population scenario produces about 1
degree less warming. (6 degrees - 1 degree by 2010 )/(7-1) =  0.83. 

So overall, even with this complicated model, incremental warming is
roughly proportional to population size.

# Capital deepening?

If population growth slows, then there will be, according to the Solow
model, capital deepening -- assuming savings rates and depreciation
stay constant. In principle, this could lessen the environmental
benefits of slower human population growth, but it's not clear that it
would be a very big effect.

How much does capital deepen in the slower population growth scenario?

To see the effect more clearly, we'll look at "k.nd" (the steady state
ignoring climate damages "D" (nd = "no damages"), which we have pretty
much ignored already.)

```{r}
k.nd = out$k.nd
k.nd.slow = out.slow.pop$k.nd
plot(year, k.nd, type = "l", main = "Capital per person, baseline and slower pop growth")
lines(year, k.nd.slow, lty = 2)
```

It looks like there's a bit of capital deepening at the beginning but
not much. How can this be?

We'll explore this in "graded questions" at the end. The answer, we'll
see, boils down to the differences in growth rates being very small
(fractions of a percent) relative to the depreciation rate (10%).

(Note: the simplified implementation of the model assumes that capital
is always at its steady state level.  One could also reprogram so that
capital is accumulated directly from the previous period, depreciated,
and recalculated per person. I don't think the answer is likely to be
very different.)

# Accelerating decarbonization

Carbon emissions are modeled as a changing fraction of economic
output. Every million dollars of output produces some number of tons
of carbon emissions. The baseline assumption that we graphed above is
that emission rates per unit of output start out declining at 1% per
year, a pace that slows very gradually later on. What if we started
out with emission decline that is twice as fast, 2%? What would
temperature be in 2100?

Let's start with a guess. 

Q3. What do we think will happen to emissions if we have much faster
   decarbonization beginning at twice the rate of baseline??

A. Even with slower emissions per unit of GDP, productivity is
   increasing so fast that increasing GDP will mean ever increasing
   emissions.

B. Over the long term, the effect of lower emissions per unit of
   output will dominate increasing output, resulting -- eventually --
   in ever declining emissions.

C. We're doomed. There's nothing we can do. And nothing matters
   anymore.

D. A & C.


```{r}
out.fast.decarb <- solow_climate_fun(gL0 = 0.022547, ## we leave starting pop growth as is
                  dL = -0.052, ## and multiply rate of change by 2
                  gsigma0 = -.01*2, dsigma = -.0002)
```

Let's plot the carbon intensity in the two scenarios just to compare

```{r}
sigma = out$sigma
sigma.fast = out.fast.decarb$sigma
## note:  here  "fast" is refering to decarbonization,
## whereas before "slow" was refering to population growth. Sorry!

plot(year, sigma, ylim = c(0,.6), main = "Carbon intensity, by scenario")
lines(year, sigma.fast)
grid()
```
It looks like emissions intensity (per $ of output) get very low by 2100 in the rapid decrease. But they're still quite far from zero well past 2050.

And let's see how total annual emissions are affected. (Remember this is emissions intensity times the total size of the economy.)

```{r}
E = out$E
E.fast = out.fast.decarb$E
plot(year, E, ylim = c(0, 25), main = "Emissions, by scenario")
lines(year, E.fast)
grid(lwd = 2)
```

We see that annual emissions rise and fall in both cases. Eventually
the decline in emissions intensity dominates the increase in
output. With faster decline in emission intensity -- our "fast"
scenario -- we get what looks to be much lower emissions. But we can
also see that even in the "fast" scenario emissions don't fall below
their current level until almost 2100.

And finally, temperature. Could decarbonization save us?

```{r}
Temp.fast = out.fast.decarb$Temp
plot(year, Temp)
lines(year, Temp.fast)
grid(lwd = 2)

```

This story seems good-news, bad-news, bad-news. The good news is that
by 2100, there would be about 1 degree less of warming -- which is a
pretty big effect. The bad news is that we would still get about the
same amount of warming by 2050.

The biggest benefit in terms of climate of faster decline emissions
intensity appears to be that it helps us avoiding the enormous
temperature increases far in the future, beyond the year 2100 into the
22nd century. If we care a lot about our great-grand-children, then
this is a very big benefit. Although the model also predicts they'll
be very rich. So,it's not entirely clear we should sacrifice
enormously now for them. (This raises the whole issue of how to
compare generations, which is very interesting but beyond the scope of
what we're looking at here.)

One more bit of optimism: a much faster rate of de-carbonization is
not implausible. Figure 6 (and Table 3) in the Nordhaus reading shows
world carbon intensity is already decreasing at something like 1.8%
per year. A combination of technological advances in energy production
and world-wide action on carbon pricing holds some hope for making
rapid progress.

Here are the take-away lessons I got from writing this lab:

* Population matters, but not as much as I expected. I think the
  reason for this is that although emissions are more-or-less
  proportional to population size in the model, population size just
  doesn't change that much if we accelerate the approach to zero
  population growth.
  
* Declines in emission intensity are potentially much more
  important. As modeled, the lower bound on sigma is zero, and the
  faster we get there, the less warming there will be.
  
* Substantial global warming is very hard to avoid. This is because we
  have a substantial amount of cumulative emissions even today, and
  they are going to continue at their present rates for some time, no
  matter what we do.
  
* Nordhaus's call to optimize and choose the best path should not be
  seen as just giving up and accepting whatever comes our way. Just
  the opposite. It is a call to act now to avoid very bad
  outcomes. But at the same time a caution that draconian reductions
  in standard of living now may not be the best thing. Instead, we
  should aim for a path that balances the costs and the benefits. His
  optimal path has about 3 degrees of warming by 2100, remarkably
  close to the one we obtained by doubling the speed of
  decarbonization.
  
# Graded questions

1. Zero population growth

a. Simulate zero population growth by modifying the "slow" scenario
   above.
   Hint 1: you need to make population growth even slower.
   (Big) Hint 2: try changing the parameters "gL" and "dL" to zero.
   Hint 3: you can check your simulation by plotting the out$L to make
   sure it is a constant.

b. How much warming will there be? Is this more or less than the 2
degrees Celsius that is considered dangerous?

2. How much decarbonization is necessary?

a. Try different rates of change of sigma (by varying the parameter
"gsigma0") until you find one that will keep world temperature rise
below 2 degrees. What value is enough? (Optional: a sentence on
whether this number seems realistic).

b. Why is Nordhaus's optimal trajectory higher than 2 degrees? (A few
sentences is fine.)

3. There will clearly be a Tragedy of the Commons if each firm is
allowed to make their own independent judgement about how much carbon
to emit. According to Nordhaus, why is it not sufficient to let each
country enforce its own carbon emissions policy on its own firms? [1
or 2 sentences is fine.]

4. For the following questions, use the "app" available 
http://shiny.demog.berkeley.edu/josh/harvest/ to gather data on
the relationship between harvest rate "h" and the sustainable
(steady-state) yield. You can do this by moving the slider from 0
upwards by increments of 0.02.

a. There is a slight bug in the app, so that it reports sustainable
yields even when harvest rate is unsustainable. How can you
inspect the graphs to check on sustainability?

b. When h = 0.01, the sustainable yield is 0.0678, what is the
sustainable yield when h = 0.02 and h = 0.04?

c. What harvest rate maximizes the sustainable yield? [Hint: try
all of the slider values]

d. What harvest rate maximizes the yield during the first year of
harvesting (this is about year 10 on the slider, or year 0 on
the time axes of the graph)? (You can either report the value
you find with the "app" or your theoretical answer)

e. Would harvest yield be higher in year "1" if we harvested at the
sustainable rate found in part 4.3, or if they were .02 higher
than the sustainable rate?


Congratulations! You've finished Lab 6. Whew!









